---
title: "Office Ratings"
author: "Pete Talbert"
date: "4/28/2020"
output: 
  word_document:
    reference_docx: "../templates/word-styles-office-2016.docx"
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 5, message = FALSE, warning = FALSE, cache = FALSE, dpi = 700)

library(tidyverse)
library(scales)
library(knitr)

theme_set(theme_minimal(base_size = 10))

office_ratings <- read_csv("data/office_ratings.csv")
```

Below I present analysis of The Office television series that ran on NBC from 2005 to 2013. The data is grouped by season and episode, with IMDB ratings, and total IMDB votes.

## How many episodes in each season?

```{r}
office_ratings %>% 
  mutate(season = as_factor(season)) %>% # SOMETHING NEW INTRODUCED
  count(season) %>% 
  ggplot(aes(x = season, y = n)) +
  geom_col(fill = "#00828A", alpha = 0.8) +
  labs(
    title = "Seasons 1 and 4 were particularly low",
    subtitle = "Count of episodes by season of The Office",
    x = "Season",
    y = "Number of episodes"
  ) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.y = element_blank())
```

Why are there only 14 episodes in season 4? According to [Wikipedia](https://en.wikipedia.org/wiki/List_of_The_Office_(American_TV_series)_episodes#Episodes), the 2007-2008 Writers Guild of America strike shut down production for nearly five months in the middle of the season. 

## What was the average IMDB rating of episodes by season?

```{r}
office_ratings %>% 
  mutate(season = as_factor(season)) %>% 
  group_by(season) %>% # SOMETHING NEW INTRODUCED
  summarize(avg_imdb_rating = mean(imdb_rating)) %>% # SOMETHING NEW INTRODUCED
  ungroup() %>% 
  ggplot(aes(x = season, y = avg_imdb_rating, group = 1)) +
  geom_line(size = 1) +
  geom_point(size = 1.5) +
  scale_y_continuous(breaks = seq(0, 10, 2), limits = c(0, 10)) +
  labs(
    title = "IMDB ratings remained fairly stead with a peak in season 4",
    subtitle = "Average IMDB rating of episodes by season of The Office",
    x = "Season",
    y = "Average IMDB rating"
  ) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.y = element_blank())
  
```

## What was the trend of IMDB ratings within a season?

```{r}
office_ratings %>% 
  mutate(season = as_factor(season)) %>% 
  ggplot(aes(x = episode, y = imdb_rating, color = season)) +
  geom_line(size = 1) + # SOMETHING NEW INTRODUCED
  geom_point(size = 1.5) +
  scale_x_continuous(breaks = seq(0, 25, 5)) +
  scale_y_continuous(breaks = seq(0, 10, 2), limits = c(0, 10)) +
  scale_color_viridis_d() +
  labs(
    title = "No real trend of IMDB rating within seasons, although\nsome do peak at the end",
    subtitle = "IMDB rating of episodes by season of The Office",
    x = "Episode no.",
    y = "IMDB rating",
    color = "Season"
  ) +
  theme(panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank())
```

## What were the top ten most popular episodes?

```{r}
office_ratings %>% 
  arrange(desc(imdb_rating)) %>% 
  head(10) %>% 
  mutate(`#` = row_number()) %>% # SOMETHING NEW INTRODUCED
  select(`#`, Season = season, Episode = episode, Title = title, `IMDB Rating` = imdb_rating) %>% 
  kable() # SOMETHING NEW INTRODUCED
```

## Is there a relationship between the IMDB rating and the number of votes?

```{r fig.height=5}
ustr::theme_set_ust_doc()

office_ratings %>% 
  ggplot(aes(x = total_votes, y = imdb_rating)) +
  geom_point() + # SOMETHING NEW INTRODUCED
  scale_x_log10(labels = comma_format()) + # SOMETHING NEW INTRODUCED
  geom_smooth(method = "loess", se = FALSE) + # SOMETHING NEW INTRODUCED
  labs(
    title = "There is a positive relationship between IMDB\nratings and total votes",
    subtitle = "IMDB ratings and total votes of The Office",
    x = "Total votes",
    y = "IMDB rating",
    caption = "Note: Line is a loess smoothing curve."
  ) +
  theme(panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank()) -> p

ggsave("office_imdb.png", p, height = 5, width = 6.5, dpi = 1000, device = "png")
```

