---
title: "College tuition, diversity, and pay data"
author: "Pete Talbert"
date: "5/8/2020"
output: word_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 5, message = FALSE, warning = FALSE, cache = FALSE, dpi = 700)

library(tidyverse)
library(scales)
library(knitr)

theme_set(theme_minimal(base_size = 10))

tuition_cost <- read_csv("data/tuition_cost.csv")

diversity_school <- read_csv("data/diversity_school.csv")

salary_potential <- read_csv("data/salary_potential.csv")


library(skimr)
```

# Skim

```{r}

tuition_cost 
summary(tuition_cost)
skim(tuition_cost)
str(tuition_cost)
glimpse(tuition_cost)

diversity_school
print(diversity_school, n = 11)
summary(diversity_school)
skim(diversity_school)
str(diversity_school)
glimpse(diversity_school)

salary_potential
summary(salary_potential)
skim(salary_potential)
str(salary_potential)
glimpse(salary_potential)

```

```{r}
print(tuition_cost, n = 15)

summary(tuition_cost)
skim(tuition_cost)

str(tuition_cost)
glimpse(tuition_cost)
```


# Tuition cost

count - fct_reorder - geom_col() - coord_flip() TECHNIQUE


```{r}
tuition_cost_um <- tuition_cost %>% 
  filter(state_code %in% c("MN", "WI", "ND", "SD", "IA"))

tuition_cost_um %>% 
  count(state_code) %>% 
  mutate(state_code = fct_reorder(state_code, n)) %>% 
  ggplot(aes(x = state_code, y = n)) +
  geom_col() +
  coord_flip()

tuition_cost_um %>% 
  count(state_code) %>% 
  mutate(state_code = fct_reorder(state_code, n)) %>% 
  ggplot(aes(x = state_code, y = n)) +
  geom_col() +
  coord_flip()

tuition_cost_um %>% 
  count(state_code, type) %>% 
  mutate(state_code = fct_reorder(state_code, n, .fun = sum)) %>% 
  ggplot(aes(x = state_code, y = n, fill = type)) +
  geom_col() +
  coord_flip()


tuition_cost_um %>% 
  count(state_code, type) %>% 
  mutate(state_code = fct_reorder(state_code, n, .fun = sum)) %>% 
  ggplot(aes(x = state_code, y = n, fill = type)) +
  geom_col(position = "fill") +
  coord_flip()

```

How would we sort highest to lowest proportion of state?

```{r}
tuition_cost_um %>% 
  count(state_code, type) %>% 
  group_by(state_code) %>% 
  mutate(pct = n / sum(n)) %>% 
  ungroup() %>% 
  mutate(public_pct = if_else(type == "Public", pct, 0),
         state_code = fct_reorder(state_code, public_pct, .fun = max)) %>% 
  ggplot(aes(x = state_code, y = n, fill = type)) +
  geom_col(position = "fill") +
  coord_flip()
```

```{r}
tuition_cost %>% 
  count(state_code, type) %>% 
  group_by(state_code) %>% 
  mutate(pct = n / sum(n)) %>% 
  ungroup() %>% 
  filter(type == "Public")
```

## Introduce `pivot_longer()` with state dataset

What if I wanted to compare the average costs by state for each type of cost column?

```{r}
tuition_cost_um %>% 
  group_by(state) %>% 
  summarize(mean(room_and_board), 
            mean(in_state_tuition), 
            mean(in_state_total), 
            mean(out_of_state_tuition), 
            mean(out_of_state_total))
```

Works fine, but how in the world would I *plot* something like this? Would have to go one at a time for each cost (because they're each in a different column).

```{r}


tuition_cost_um %>% 
  pivot_longer(cols = c(room_and_board, in_state_tuition, in_state_total, out_of_state_tuition, out_of_state_total), 
               names_to = "cost_type", 
               values_to = "cost") %>% filter(str_detect(name, "Iowa"))
  group_by(state, cost_type) %>% 
  summarize(avg_cost = mean(cost))
```

Gotta deal with those NAs.

## Introduce `pivot_wider()` with the diversity dataset

```{r}
diversity_school %>% 
  pivot_wider(names_from = category, values_from = enrollment)
# error: explain
# filter NA college data
diversity_school %>% 
  pivot_wider(names_from = category, values_from = enrollment, values_fn = length) %>% View()

diversity_school_non_na <- diversity_school %>% 
  filter(!is.na(name))

diversity_school_non_na %>% 
  pivot_wider(names_from = category, 
              values_from = enrollment) %>% 
  print(., width = 300)


```

## Presentation-ready plots

```{r}
tuition_cost %>% 
  filter(state_code %in% c("MN", "WI", "ND", "SD", "IA")) %>% 
  ggplot(aes(x = state, y = in_state_tuition, color = state)) +
  geom_jitter(width = 0.05) +
  scale_y_continuous(limits = c(0, NA), labels = dollar_format(), breaks = c(0, 10000, 20000, 30000, 40000, 50000, 60000)) +
  scale_color_brewer(palette = 10)

p <- tuition_cost %>% 
  filter(state_code %in% c("MN", "WI", "ND", "SD", "IA")) %>% 
  group_by(state_code, type) %>% 
  summarize(avg_in_state_tuition = mean(in_state_tuition)) %>% 
  ungroup() %>% 
  ggplot(aes(x = state_code, y = avg_in_state_tuition, fill = state_code)) +
  geom_col() +
  geom_text(aes(label = dollar(avg_in_state_tuition, accuracy = 1)), size = 3, vjust = 1.1) +
  scale_y_continuous(limits = c(0, NA), labels = dollar_format(), breaks = c(0, 10000, 20000, 30000, 40000, 50000, 60000)) +
  scale_fill_brewer(palette = 16, type = "seq") + 
  facet_wrap(~ type, nrow = 2) +
  theme_minimal() +
  theme(legend.position = "none", 
        panel.grid.major.x = element_blank(), 
        panel.grid.minor.y = element_blank(), plot.title = element_text(face = "bold")) +
  labs(
    title = "MN has most expensive private universities\nin the upper-midwest",
    subtitle = "Avg in-state tuition by type",
    x = element_blank(),
    y = element_blank()
  )

p

ggsave("um-tuition.png", p, width = 6, height = 4, units = "in", dpi = 700)

```


```{r}
tuition_cost %>% 
  filter(state_code == "MN", degree_length == "4 Year", !is.na(room_and_board)) %>% 
  pivot_longer(cols = c(room_and_board, in_state_tuition, in_state_total, out_of_state_tuition, out_of_state_total), 
               names_to = "cost_type", 
               values_to = "cost") %>% 
  mutate(cost_type = str_to_sentence(str_replace_all(cost_type, "_", " "))) %>% 
  group_by(type, cost_type) %>% 
  summarize(avg_cost = mean(cost)) %>% 
  ggplot(aes(x = cost_type, y = avg_cost, fill = type)) +
  geom_col(position = "dodge") +
  coord_flip()
```




## Introduce `inner_join()` with state dataset


```{r}
states <- bind_cols(state_code = state.abb, state_region = state.region, state_division = state.division)


tuition_cost_region <- tuition_cost %>% 
  inner_join(states, by = "state_code") %>% 
  select(name, state, state_code, state_region, state_division, room_and_board, in_state_tuition)
```

Perfect example of inner vs left join!

```{r}
tuition_cost_region %>% 
  count(state_region, state_code, type) %>% 
  group_by(state_code) %>% 
  mutate(pct = n / sum(n)) %>% 
  ungroup() %>% 
  mutate(public_pct = if_else(type == "Public", pct, 0),
         state_code = fct_reorder(state_code, public_pct, .fun = max)) %>% 
  ggplot(aes(x = state_code, y = n, fill = type)) +
  geom_col(position = "fill") +
  coord_flip() +
  facet_wrap(~ state_region, scales = "free_y")

tuition_cost_region %>% 
  count(state_division, state_code, type) %>% 
  group_by(state_code) %>% 
  mutate(pct = n / sum(n)) %>% 
  ungroup() %>% 
  mutate(public_pct = if_else(type == "Public", pct, 0),
         state_code = fct_reorder(state_code, public_pct, .fun = max)) %>% 
  ggplot(aes(x = state_code, y = n, fill = type)) +
  geom_col(position = "fill") +
  coord_flip() +
  facet_wrap(~ state_division, scales = "free_y")

tuition_cost_region %>% 
  count(state_division, state_code, type) %>% 
  group_by(state_code) %>% 
  mutate(pct = n / sum(n)) %>% 
  ungroup() %>% 
  mutate(public_pct = if_else(type == "Public", pct, 0), # NEED TO FIND A BETTER WAY TO ILLUSTRATE THIS.
         state_code = fct_reorder(state_code, public_pct, .fun = max)) %>% 
  ggplot(aes(x = state_code, y = n, fill = type)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ state_division, scales = "free_y")
```


NA values for state!!!

## `group_by()`, `mutate()`, and window functions

```{r}
tuition_cost_um %>% 
  mutate(cost_rank = min_rank(desc(in_state_tuition))) %>% 
  select(name, state, type, in_state_tuition, cost_rank)
```


## Let's hone in on the upper midwest just to make things a little easier

```{r}
tuition_cost_um <- tuition_cost %>% 
  filter(state_code %in% c("MN", "WI", "ND", "SD", "IA"))

tuition_cost_um %>% 
  count(state_code)

tuition_cost_um %>% 
  count(state_code, type)

tuition_cost_um %>% 
  count(state_code, type, degree_length)

```

## Who's the most expensive in the upper midwest?

```{r}
tuition_cost_um %>% 
  arrange(desc(out_of_state_total)) %>% 
  head(10)
```

## Relationship between room and board and tuition?

```{r}
tuition_cost %>% 
  ggplot(aes(x = room_and_board, y = out_of_state_tuition)) +
  geom_point()
```


```{r}
tuition_cost %>% 
  ggplot(aes(x = room_and_board, y = out_of_state_tuition, color = type)) +
  geom_point()

tuition_cost %>% 
  ggplot(aes(x = room_and_board, y = out_of_state_tuition, color = type)) +
  geom_point() +
  scale_y_log10()

tuition_cost %>% 
  ggplot(aes(x = room_and_board, y = out_of_state_tuition, color = type)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

Wow, seems like although private universities are obviously more expensive, their room and board costs may be relatively cheaper. How would we try to determine that?

```{r}
# more uses for mutate

tuition_cost %>% 
  mutate(upper_mid = if_else(state_code %in% c("MN", "WI", "ND", "SD", "IA"), "Y", "N"),
         room_tuition_ratio = room_and_board / out_of_state_tuition) %>% 
  ggplot(aes(x = room_tuition_ratio, fill = type)) +
  geom_histogram(binwidth = 0.1)


tuition_cost %>% 
  mutate(upper_mid = if_else(state_code %in% c("MN", "WI", "ND", "SD", "IA"), "Y", "N"),
         room_tuition_ratio = room_and_board / out_of_state_tuition) %>% 
  #filter(room_tuition_ratio < 3) %>% 
  ggplot(aes(x = room_tuition_ratio, fill = type)) +
  geom_histogram(binwidth = 0.1) +
  scale_x_log10()

tuition_cost %>% 
  mutate(upper_mid = if_else(state_code %in% c("MN", "WI", "ND", "SD", "IA"), "Y", "N"),
         room_tuition_ratio = room_and_board / out_of_state_tuition) %>% 
  filter(room_tuition_ratio < 2) %>% 
  ggplot(aes(x = room_tuition_ratio, fill = type)) +
  geom_histogram(binwidth = 0.1)

# what happens when we have NA values????

tuition_cost %>% 
  mutate(upper_mid = if_else(state_code %in% c("MN", "WI", "ND", "SD", "IA"), "Y", "N"),
         room_tuition_ratio = room_and_board / out_of_state_tuition) %>% 
  group_by(type) %>% 
  summarize(mean_ratio = mean(room_tuition_ratio),
            median_ratio = median(room_tuition_ratio))

tuition_cost %>% 
  mutate(upper_mid = if_else(state_code %in% c("MN", "WI", "ND", "SD", "IA"), "Y", "N"),
         room_tuition_ratio = room_and_board / out_of_state_tuition) %>% 
  filter(!is.na(room_tuition_ratio)) %>% 
  group_by(type) %>% 
  summarize(mean_ratio = mean(room_tuition_ratio),
            median_ratio = median(room_tuition_ratio),
            n = n())
```

This shows that on average, room and board is relatively cheaper at private universities...

```{r}
tuition_cost %>% 
  mutate(upper_mid = if_else(state_code %in% c("MN", "WI", "ND", "SD", "IA"), "Y", "N"),
         room_tuition_ratio = room_and_board / out_of_state_tuition) %>% 
  ggplot(aes(x = room_tuition_ratio, y = out_of_state_tuition)) + 
  geom_point()

#mega transformed!!!
tuition_cost %>% 
  mutate(upper_mid = if_else(state_code %in% c("MN", "WI", "ND", "SD", "IA"), "Y", "N"),
         room_tuition_ratio = room_and_board / out_of_state_tuition) %>% 
  filter(room_tuition_ratio > 0.1, type != "For Profit") %>% 
  ggplot(aes(x = room_tuition_ratio, y = out_of_state_tuition)) + 
  geom_point(aes(color = type), alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_smooth(method = "loess", se = FALSE) +
  scale_x_log10() +
  scale_y_log10(labels = comma_format())
```

It seems like universities who chargle less for tuition tend to have to charge more for room and board to make up the difference???

# What does the public-private breakdown look like by state?

```{r}
tuition_cost_pp <- tuition_cost %>% 
  filter(type %in% c("Public", "Private"))

tuition_cost_pp %>% 
  count(state_code, type) %>% 
  group_by(state_code) %>% 
  mutate(pct = n / sum(n)) %>% 
  filter(type == "Public") %>% 
  ggplot(aes(x = state_code, y = pct)) +
  geom_col() +
  coord_flip()
```

